Evolution of the AGN UV luminosity function from redshift 7.5

### 2018-07-23 Paper/paper.tex

\begin{abstract}
  Determinations of the UV luminosity function of AGN at high
  redshifts are important for constraining the AGN contribution to
  reionization and understanding the growth of supermassive black
  holes.  Recent inferences of the luminosity function suffer from
  inconsistencies arising from inhomogeneous selection and mixing of
  AGN data.  We address this problem by constructing a sample of more
  than 80,000 colour-selected AGN from redshift $z=0$ to $7.5$.  While
  this sample is composed of multiple data sets with spectroscopic
  redshifts and completeness estimates, we homogenise these data sets
  to identical cosmologies, intrinsic AGN spectra, and magnitude
  systems.  Using this sample, we derive the AGN UV luminosity
  function from redshifts $z=0$ to $7.5$.  The luminosity function has
  a double power law form at $z<5.5$, and a single power law form at
  higher redshifts.  The break luminosity brightens rapidly with
  redshift and the faint end of the luminosity function steepens.  In
  spite of this steepening, the contribution of AGN to the hydrogen
  photoionization rate at $z\sim 6$ is subdominant, although it can be
  non-negligible if these luminosity functions hold down to
  $M_{1450}=-18$.  Under reasonable assumptions, AGN can reionize
  \HeII\ by redshift $z=3.2$.  At low redshifts ($z<0.5$), AGN can
  produce all of the hydrogen photoionization rate inferred from the
  statistics of \HI\ absorption lines in the IGM.  Our global analysis
  of the luminosity function also reveals important systematic errors
  in the data, particularly at $z=2.2$--$3.5$, which need to be
  addressed in future in order to improve our results.  We make
  various fitting functions, luminosity function analysis codes, and
  homogenised AGN data publicly available.
\end{abstract}

@@

The luminosity function of active galactic nuclei (AGN) and its
evolution over cosmological time scales has been a matter of central
interest of a large body of work over the last four decades
\citep{1968ApJ...151..393S, 1978A&A....68...17M, 1983ApJ...269..352S,
  1988ApJ...325...92K, 1988MNRAS.235..935B, 1993ApJ...406L..43H,
  1994ApJ...421..412W, 1995AJ....110...68S, 1995AJ....110.2553K,
  1995ApJ...438..623P, 2000MNRAS.317.1014B, 2001AJ....121...54F,
  2004AJ....128..515F, 2006AJ....131.2766R, 2007ApJ...654..731H,
  2009MNRAS.392...19C, 2010AJ....139..906W, 2011ApJ...728L..26G,
  2013ApJ...773...14R, 2013ApJ...768..105M, 2015AA...578A..83G,
  2015ApJ...798...28K, 2016ApJ...829...33Y, 2016ApJ...833..222J}.
Determination of the AGN luminosity function constrains models of the
build-up of supermassive black holes \citep{1982MNRAS.200..115S,
  2004ApJ...602..603Y, 2007ApJ...654..731H, 2008ApJ...679..118S,
  2009A&A...493...55E, 2000ApJ...531...42H, 2010MNRAS.401.2531A,
  2013ApJ...773...14R, 2014ApJ...787...73D, 2015MNRAS.448.3603D,
  2015MNRAS.452..575S, 2016MNRAS.462..190R}.  Due to the incidence of
supermassive black holes in most galaxies, the tight scaling relations
observed to exist between the mass of these black holes and properties
of their host galaxies, and the increasing consensus that AGN activity
feeds back on the host galaxy evolution, the AGN luminosity function
also constrains models of galaxy formation \citep{2006ApJ...650...42L,
  2008MNRAS.385.1846M}.  Finally, thanks to their relative brightness
and high Lyman-continuum (LyC) escape fractions, the luminosity
function of AGN determines their contribution to the radiation
background that influences the temperature and ionization state of the
intergalactic medium (IGM), possibly even primarily driving hydrogen
and helium reionization in the last twelve billion years
\citep{2012ApJ...746..125H, 2015AA...578A..83G, 2017ApJ...847L..15O,
  2018MNRAS.474.2904P, 2018arXiv180104931P}.

There is a renewed interest in understanding the evolution of the UV
luminosity function of AGN, triggered by the discovery of 19
low-luminosity ($-18.9>M_{1450}>-22.6$) AGN candidates between
redshifts $z=4.1$ and $6.3$ by \citet{2015AA...578A..83G} using a
novel X-ray/NIR selection criterion.  This finding suggested that if
100\% LyC escape fraction is assumed AGN brighter than $M_{1450}=-18$
can potentially produce all of the metagalactic hydrogen
photoionization rate inferred from the \lya forest at $4<z<6$, despite
the fact that these AGN are fainter than the brightest star-forming
galaxies at these redshifts.  (The UV luminosity function of galaxies
at $z=6$ has $M^*_{1450}\sim -20$.)  A significant presence of AGN at
high redshift ($z\sim 6$) and a dominant contribution of AGN to
reionization is appealing as the LyC escape fraction of galaxies is
uncertain.  High-redshift galaxies down to rest-frame UV magnitude
$M_\mathrm{UV}=-12.5$ ($L\sim 10^{-3}L^*_\mathrm{galaxies}$) at $z=6$
\citep{2017ApJ...835..113L} and redshifts up to $z=11.1$
\citep{2016ApJ...819..129O} have now been reported.  But the escape of
LyC photons has been measured in only a few comparatively bright
($L>0.5L^*_\mathrm{galaxies}$) galaxies at relatively low redshifts
($z < 4$).  The escape fraction in these galaxies reveals a broad
distribution from less than 2\% to more than 80\%
\citep{2010ApJ...725.1011V, 2011ApJ...736...41B, 2015ApJ...804...17S,
  2015ApJ...810..107M, 2016A&A...585A..48G, 2017MNRAS.468..389J,
  2018arXiv180506071S, 2018arXiv180601741F, 2018arXiv180511621M,
  2018arXiv180303655C}, but the average escape fraction is typically
lower than 20\%.  Statistical constraints from \HI\ column density
measurements in gamma-ray burst (GRB) afterglow spectra suggest an
even lower escape fraction of 0.5\% \citep{2007ApJ...667L.125C,
  2009ApJS..185..526F, 2018arXiv180507318T}.  This is a challenge for
reionization models, which require an escape fraction of about 20\% in
galaxies as faint as $M_\mathrm{UV}=-13$ \citep{2016PASA...33...37F,
  2015ApJ...802L..19R, 2016MNRAS.457.4051K}.  An enhanced incidence of
high-redshift AGN may also be consistent with the shallow bright-end
slopes of the $z\sim 7$ UV luminosity function of galaxies relative to
a Schechter function \citep{2012MNRAS.426.2772B, 2014MNRAS.440.2810B,
  2014ApJ...792...76B, 2015MNRAS.452.1817B} and the hard spectra of
these galaxies \citep{2015MNRAS.450.1846S, 2015MNRAS.454.1393S,
  2017MNRAS.464..469S}.  Finally, AGN may also provide a natural
explanation for the large scatter in the \lya opacity between
different quasar sightlines close to redshift $z=6$
\citep{2015MNRAS.447.3402B, 2018arXiv180208177B}.
These opacity fluctuations extend to substantially larger scales
($\gtrsim 50\, h^{-1}$cMpc) than expected in galaxies-dominated
reionization models (\citealt{2015MNRAS.453.2943C}; although see
\citealt{2016MNRAS.460.1328D, 2015ApJ...813L..38D,
  2018arXiv180308932B}).  AGN clustering can result in these
fluctuations naturally if there is a significant contribution
($\gtrsim 50\%$) of AGN to the ionising emissivity at $z=5$--$6$
\citep{2017MNRAS.465.3429C}.

@@

We included surveys based on a set of simple criteria:
\begin{enumerate}
\item High spectroscopic completeness of the target sample.
\item Accurate rest-frame UV-optical CCD photometry.
\item Statistical power (sample size, coverage in $z$ and/or absolute magnitude).
\item A well-characterised selection function.
\end{enumerate}
As a prerequisite for a joint analysis of the QLF we obtained the
survey selection functions in electronic form, either from the
publication or by request from the authors. As a reference for future
surveys we make them electronically available here in modified and
homogenised form (see Section~\ref{sect:datahom} and
Appendix~\ref{sec:code}).

@@

At $4.7\le z<5.5$ we combined several recent surveys, accounting for
sample overlap and updated selection functions. At the bright end of
the QLF we used the 99 quasars from the SDSS+WISE survey
\citep{2016ApJ...829...33Y} that have $M_{1450}\le -26.73$ in our adopted
cosmology.  For these 99 quasars selected from 14,555\,deg$^2$ we
adopted the \citet{2016ApJ...829...33Y} selection function.  The
\citet{2016ApJ...829...33Y} sample partially overlaps with the SDSS
DR7 sample from \citet{2013ApJ...768..105M}, so to avoid
double-counting quasars we used the latter sample only at
$M_{1450}>-26.73$, yielding 103 additional $4.7\le z<5.5$ quasars selected in
6248\,deg$^2$. We used the $z\sim 5$ SDSS DR7 selection function from
\citet{2013ApJ...768..105M} that supersedes the one from
\citet{2006AJ....131.2766R} due to improved bandpass corrections and IGM
parameterization.  To these two bright-end samples we added the
faint-end sample from the \citet{2013ApJ...768..105M} SDSS Stripe~82
survey (59 uniformly selected $M_{1450}>-26.73$ $4.7\le z<5.5$ quasars in 235\,deg$^2$)
and two $4.7\le z<5.5$ quasars from \citet{2011ApJ...728L..26G},
adopting the respective selection functions.  We did not consider the
limit on the $z\sim 5$ QLF by \citet{2012ApJ...756..160I} due to
systematic errors in their selection
function\footnote{\citet{2012ApJ...756..160I} underestimated the
  dispersion in rest-frame UV quasar colours with respect to SDSS at
  all redshifts (their Figure~4). Contrary to their claim, photometric
  errors have a small effect on the colour distribution of SDSS
  quasars given the statistical errors of $<0.03$\,mag in $gri$ for 90
  per cent of the SDSS DR7 bright quasar sample ($i<19.1$) and a
  relative calibration error of $\sim 1$ per cent
  \citep{2008ApJ...674.1217P}.}.

@@

\begin{figure*}
  \begin{center}
    % rtg2.draw_g_paper() 
    \includegraphics[scale=0.65]{g.pdf}
  \end{center}
  \caption{AGN contribution to the hydrogen photoionisation rate,
    assuming unit escape fraction, when the AGN luminosity function is
    integrated down to $M_{1450}=-21$ (blue curve and shaded region)
    and $M_{1450}=-18$ (red curve and shaded region).  The shaded
    regions show the one-sigma (68.26\%) uncertainty.  Also shown are
    the photoionization rate measurements by \citet[filled
      circles]{2013MNRAS.436.1023B}, \citet[inverted
      triangles]{2011MNRAS.412.2543C}, and
    \citet[pentagons]{2017MNRAS.467.3172G}, and models of
    \citet[dotted brown curve]{2012ApJ...746..125H}, the QSO
    contribution in this model (dashed grey), \citet[dashed
      brown]{2015ApJ...813L...8M}, the QSO contribution from the model
    of \citet[dashed orange]{2015MNRAS.451L..30K}, \citet[dotted
      grey]{2017ApJ...837..106O}, and \citet[dashed
      grey]{2018arXiv180104931P}.}
  \label{fig:gammapi}
\end{figure*}

@@

Our calculation of the AGN contribution to the UV background follows
previous work on UV background synthesis models
\citep[e.g.][]{1996ApJ...461...20H,2012ApJ...746..125H}.  The main
quantity of interest is the \ion{H}{i} photoionization rate of the UV
background
\begin{equation}
  \Gamma_\ion{H}{i}\left(z\right)=\int_{\nu_{912}}^\infty\mathrm{d}\nu
  \frac{4\pi J_\nu\left(\nu,z\right)}{h\nu} \sigma_\ion{H}{i}\left(\nu\right)\quad,
\end{equation}
where $h$ is Planck's constant, $\sigma_\ion{H}{i}(\nu)$ is the
\ion{H}{i} photoionization cross-section \citep{1997MNRAS.292...27H,
  2011piim.book.....D}, and
\begin{multline}
  J_\nu(\nu, z)=\frac{c}{4\pi}\int_{z}^\infty\mathrm{d}z^\prime\frac{\left(1+z\right)^3}{H\left(z^\prime\right)\left(1+z^\prime\right)}\epsilon_\nu\left(\nu_\mathrm{em},z^\prime\right)\\
  \times\exp{\left[-\tau_\mathrm{eff}\left(\nu, z, z^\prime\right)\right]}
  \label{eqn:flux}
\end{multline}
is the angle- and space-averaged specific intensity of the UV
background. In the above equation
$H\left(z^\prime\right)=H_0(\Omega_\mathrm{m}\left(1+z^\prime\right)^3+\Omega_\Lambda)^{1/2}$
is the Hubble parameter, and
$\epsilon_\nu\left(\nu_\mathrm{em},z^\prime\right)$ is the comoving
emissivity of all \ion{H}{i} Lyman continuum sources at redshift
$z^\prime>z$ and emitted frequency
$\nu_\mathrm{em}=\nu\left(1+z^\prime\right)/\left(1+z\right)>\nu_{912}$.
In practice, we adopt an upper limit $z^\prime_\mathrm{max}=15$ when
integrating Equation~\eqref{eqn:flux}.  Assuming that quasars are the
only ionizing sources and considering Equation~\eqref{eqn:sed} we have
\begin{equation}
  \epsilon_\nu\left(\nu_\mathrm{em},z^\prime\right) = \epsilon_{912}\left(z^\prime\right)\left(\frac{\nu_\mathrm{em}}{\nu_{912}}\right)^{-1.70}\quad,
  \label{eqn:epsilon_freq}
\end{equation}
with $\epsilon_{912}\left(z^\prime\right)$ being represented by
Equations~\eqref{eqn:e912_18} and \eqref{eqn:e912_21} for our two
magnitude limits, respectively.  For Poisson-distributed absorbers
with an \ion{H}{i} column density distribution
$f\left(N_\ion{H}{i},z^{\prime\prime}\right)=\partial^2n/\left(\partial
N_\ion{H}{i}\partial z^{\prime\prime}\right)$, the effective optical
depth to \ion{H}{i} Lyman continuum photons travelling between
redshifts $z^\prime$ and $z$ is \citep{1980ApJ...240..387P}
\begin{equation}
  \tau_\mathrm{eff}\left(\nu,z,z^\prime\right) = \int_{z}^{z^\prime}\mathrm{d}z^{\prime\prime}\int_0^\infty
  \mathrm{d}N_\ion{H}{i} f\left(N_\ion{H}{i},z^{\prime\prime}\right)\left[1-e^{-\tau_1}\right]\quad,
   \label{eqn:taueff}
\end{equation}
where $\tau_1\approx
N_\ion{H}{i}\sigma_\ion{H}{i}\nu\left(1+z^{\prime\prime}\right)/\left(1+z\right)$
is the Lyman continuum optical depth through an individual
absorber\footnote{The helium content of the absorber can be ignored
  due to the small \ion{H}{i} photoionization cross section at
  $\lambda<304$\,\AA.}.  For the \ion{H}{i} column density
distribution $f\left(N_\ion{H}{i},z\right)$ we adopt the piecewise
power-law parametrization by \citet{2012ApJ...746..125H} that is
consistent with $f\left(N_\ion{H}{i},z\right)$ measurements at
$z<3.5$, and roughly reproduces both the \ion{H}{i} Ly$\alpha$
effective optical depth \citep[but not in detail---see][]{2015MNRAS.450.4081P,2017MNRAS.464..897B,2017ApJ...837..106O}
and the measured mean free path to \ion{H}{i} Lyman limit photons at
$z<5.5$ \citep{2009ApJ...705L.113P,2014MNRAS.445.1745W}.
\citet{2018arXiv180109693K} suggest that variations in
$f\left(N_\ion{H}{i},z\right)$ result in modest (10--40 per cent)
changes in $\Gamma_\ion{H}{i}$ at $z<3$. We note, however, that at
$z\ga 3.5$ all UV background synthesis models are based on brazen
extrapolations of $f\left(N_\ion{H}{i},z\right)$, whose detailed shape
for (partial) Lyman limit systems is not well constrained at these
redshifts \citep{2010ApJ...718..392P}.

@@

\begin{figure*}
  \begin{center}
    % corner.corner() with bins.lfs[-4]
    \includegraphics[width=0.6\textwidth]{corner_z3p9.pdf}
  \end{center}
  \caption{Posterior distributions of the four double-power-law
    parameters in the $3.7\leq z < 4.1$ redshift bin.  The blue
    squares indicate median values.  Similar behaviour of the
    posterior distributions is seen in all other redshift bins defined
    in Section~\ref{sec:bins}.
    \label{fig:corner}}
\end{figure*}

@@

    \textbf{GW: Delete the zbin column that we do not use. Indicate the credible (fitted) redshift bins!}

@@

  \begin{tabular}{cccc....}

@@

    $z_\mathrm{bin}$ &

@@

    0.31 & 0.25 & 0.10 & 0.40 & 0.53^{+0.02}_{-0.02} & 0.71^{+0.03}_{-0.03} & 0.30^{+0.01}_{-0.01} & 0.40^{+0.01}_{-0.01} \\
    0.50 & 0.50 & 0.40 & 0.60 & 0.75^{+0.01}_{-0.01} & 1.00^{+0.02}_{-0.02} & 0.58^{+0.01}_{-0.01} & 0.78^{+0.01}_{-0.01} \\
    0.72 & 0.70 & 0.60 & 0.80 & 1.73^{+0.06}_{-0.05} & 2.31^{+0.07}_{-0.06} & 1.19^{+0.02}_{-0.02} & 1.57^{+0.02}_{-0.02} \\
    0.91 & 0.90 & 0.80 & 1.00 & 2.85^{+0.16}_{-0.16} & 3.77^{+0.20}_{-0.19} & 2.10^{+0.05}_{-0.05} & 2.79^{+0.06}_{-0.06} \\
    1.10 & 1.10 & 1.00 & 1.20 & 3.71^{+0.11}_{-0.10} & 4.94^{+0.15}_{-0.13} & 2.91^{+0.05}_{-0.05} & 3.87^{+0.06}_{-0.06} \\

>>>>

\begin{abstract}
  Determinations of the UV luminosity function of AGN at high
  redshifts are important for constraining the AGN contribution to
  reionization and understanding the growth of supermassive black
  holes.  Recent inferences of the luminosity function suffer from
  inconsistencies arising from inhomogeneous selection and mixing of
  AGN data.  We address this problem by constructing a sample of more
  than 80,000 colour-selected AGN from redshift $z=0$ to $7.5$.  While
  this sample is composed of multiple data sets with spectroscopic
  redshifts and completeness estimates, we homogenise these data sets
  to identical cosmologies, intrinsic AGN spectra, and magnitude
  systems.  Using this sample, we derive the AGN UV luminosity
  function from redshifts $z=0$ to $7.5$.  The luminosity function has
  a double power law form at most redshifts.  The break luminosity
  brightens rapidly with redshift and the faint end of the luminosity
  function steepens.  In spite of this steepening, the contribution of
  AGN to the hydrogen photoionization rate at $z\sim 6$ is
  subdominant, although it can be non-negligible if these luminosity
  functions hold down to $M_{1450}=-18$.  Under reasonable
  assumptions, AGN can reionize \HeII\ by redshift $z=2.9$.  At low
  redshifts ($z<0.5$), AGN can produce about half of the hydrogen
  photoionization rate inferred from the statistics of \HI\ absorption
  lines in the IGM.  Our global analysis of the luminosity function
  also reveals important systematic errors in the data, particularly
  at $z=2.2$--$3.5$, which need to be addressed in future in order to
  improve our results.  We make various fitting functions, luminosity
  function analysis codes, and homogenised AGN data publicly
  available.
\end{abstract}

@@

The luminosity function of active galactic nuclei (AGN) and its
evolution over cosmological time scales has been a matter of central
interest of a large body of work over the last five decades
\citep{1968ApJ...151..393S, 1978A&A....68...17M, 1983ApJ...269..352S,
  1988ApJ...325...92K, 1988MNRAS.235..935B, 1993ApJ...406L..43H,
  1994ApJ...421..412W, 1995AJ....110...68S, 1995AJ....110.2553K,
  1995ApJ...438..623P, 2000MNRAS.317.1014B, 2001AJ....121...54F,
  2004AJ....128..515F, 2006AJ....131.2766R, 2007ApJ...654..731H,
  2009MNRAS.392...19C, 2010AJ....139..906W, 2011ApJ...728L..26G,
  2013ApJ...773...14R, 2013ApJ...768..105M, 2015AA...578A..83G,
  2015ApJ...798...28K, 2016ApJ...829...33Y, 2016ApJ...833..222J}.
Determination of the AGN luminosity function constrains models of the
build-up of supermassive black holes \citep{1982MNRAS.200..115S,
  2004ApJ...602..603Y, 2007ApJ...654..731H, 2008ApJ...679..118S,
  2009A&A...493...55E, 2000ApJ...531...42H, 2010MNRAS.401.2531A,
  2013ApJ...773...14R, 2014ApJ...787...73D, 2015MNRAS.448.3603D,
  2015MNRAS.452..575S, 2016MNRAS.462..190R}.  Due to the incidence of
supermassive black holes in most galaxies, the tight scaling relations
observed to exist between the mass of these black holes and properties
of their host galaxies, and the increasing consensus that AGN activity
feeds back on the host galaxy evolution, the AGN luminosity function
also constrains models of galaxy formation \citep{2006ApJ...650...42L,
  2008MNRAS.385.1846M}.  Finally, thanks to their relative brightness
and high Lyman-continuum (LyC) escape fractions, the luminosity
function of AGN determines their contribution to the radiation
background that influences the temperature and ionization state of the
intergalactic medium (IGM), possibly even primarily driving hydrogen
and helium reionization in the last twelve billion years
\citep{2012ApJ...746..125H, 2015AA...578A..83G, 2017ApJ...847L..15O,
  2018MNRAS.474.2904P, 2018arXiv180104931P}.

There is a renewed interest in understanding the evolution of the UV
luminosity function of AGN, triggered by the discovery of 19
low-luminosity ($-18.9>M_{1450}>-22.6$) AGN candidates between
redshifts $z=4.1$ and $6.3$ by \citet{2015AA...578A..83G} using a
novel X-ray/NIR selection criterion.  This finding suggested that if
100\% LyC escape fraction is assumed AGN brighter than $M_{1450}=-18$
can potentially produce all of the metagalactic hydrogen
photoionization rate inferred from the \lya forest at $4<z<6$, despite
the fact that these AGN are fainter than the brightest star-forming
galaxies at these redshifts.  (The UV luminosity function of galaxies
at $z=6$ has characteristic magnitude $M^*_{1450}\sim -20$.)  A
significant presence of AGN at high redshift ($z\sim 6$) and a
dominant contribution of AGN to reionization is appealing as the LyC
escape fraction of galaxies is uncertain.  High-redshift galaxies down
to rest-frame UV magnitude $M_\mathrm{UV}=-12.5$ ($L\sim
10^{-3}L^*_\mathrm{galaxies}$) at $z=6$ \citep{2017ApJ...835..113L}
and redshifts up to $z=11.1$ \citep{2016ApJ...819..129O} have now been
reported.  But the escape of LyC photons has been measured in only a
few comparatively bright ($L>0.5L^*_\mathrm{galaxies}$) galaxies at
relatively low redshifts ($z < 4$).  The escape fraction in these
galaxies reveals a broad distribution from less than 2\% to more than
80\% \citep{2010ApJ...725.1011V, 2011ApJ...736...41B,
  2015ApJ...804...17S, 2015ApJ...810..107M, 2016A&A...585A..48G,
  2017MNRAS.468..389J, 2018arXiv180506071S, 2018arXiv180601741F,
  2018arXiv180511621M, 2018arXiv180303655C}, but the average escape
fraction is typically lower than 20\%.  Statistical constraints from
\HI\ column density measurements in gamma-ray burst (GRB) afterglow
spectra suggest an even lower escape fraction of 0.5\%
\citep{2007ApJ...667L.125C, 2009ApJS..185..526F, 2018arXiv180507318T}.
This is a challenge for reionization models, which require an escape
fraction of about 20\% in galaxies as faint as $M_\mathrm{UV}=-13$
\citep{2016PASA...33...37F, 2015ApJ...802L..19R, 2016MNRAS.457.4051K}.
An enhanced incidence of high-redshift AGN may also be consistent with
the shallow bright-end slopes of the $z\sim 7$ UV luminosity function
of galaxies relative to a Schechter function
\citep{2012MNRAS.426.2772B, 2014MNRAS.440.2810B, 2014ApJ...792...76B,
  2015MNRAS.452.1817B} and the hard spectra of these galaxies
\citep{2015MNRAS.450.1846S, 2015MNRAS.454.1393S, 2017MNRAS.464..469S}.
Finally, AGN may also provide a natural explanation for the large
scatter in the \lya opacity between different quasar sightlines close
to redshift $z=6$ \citep{2015MNRAS.447.3402B, 2018arXiv180208177B}.
These opacity fluctuations extend to substantially larger scales
($\gtrsim 50\, h^{-1}$cMpc) than expected in galaxies-dominated
reionization models (\citealt{2015MNRAS.453.2943C}; although see
\citealt{2016MNRAS.460.1328D, 2015ApJ...813L..38D,
  2018arXiv180308932B}).  AGN clustering can result in these
fluctuations naturally if there is a significant contribution
($\gtrsim 50\%$) of AGN to the ionising emissivity at $z=5$--$6$
\citep{2017MNRAS.465.3429C}.

@@

We included surveys based on a set of simple criteria:
\begin{enumerate}
\item High spectroscopic completeness of the target sample.
\item Accurate rest-frame UV-optical CCD photometry.
\item Statistical power (sample size, coverage in $z$ and/or absolute magnitude).
\item A well-characterised selection function.
\end{enumerate}
As a prerequisite for a joint analysis of the QSO luminosity function
(QLF) we obtained the survey selection functions in electronic form,
either from the publication or by request from the authors. As a
reference for future surveys we make them electronically available
here in modified and homogenised form (see Section~\ref{sect:datahom}
and Appendix~\ref{sec:code}).

@@

At $4.7\le z<5.5$ we combined several recent surveys, accounting for
sample overlap and updated selection functions. At the bright end of
the QLF we used the 99 quasars from the SDSS+WISE survey
\citep{2016ApJ...829...33Y} that have $M_{1450}\le -26.73$ in our
adopted cosmology.  For these 99 quasars selected from 14,555\,deg$^2$
we adopted the \citet{2016ApJ...829...33Y} selection function.  The
\citet{2016ApJ...829...33Y} sample partially overlaps with the
\citet{2013ApJ...768..105M} SDSS DR7 sample, so to avoid
double-counting quasars we used the latter sample only at
$M_{1450}>-26.73$, yielding 103 additional $4.7\le z<5.5$ quasars
selected in 6248\,deg$^2$. We used the $z\sim 5$ SDSS DR7 selection
function from \citet{2013ApJ...768..105M} that supersedes the one from
\citet{2006AJ....131.2766R} due to improved bandpass corrections and
IGM parameterization.  To these two bright-end samples we added the
faint-end sample from the \citet{2013ApJ...768..105M} SDSS Stripe~82
survey (59 uniformly selected $M_{1450}>-26.73$ $4.7\le z<5.5$ quasars
in 235\,deg$^2$) and two $4.7\le z<5.5$ quasars from
\citet{2011ApJ...728L..26G}, adopting the respective selection
functions.  We did not consider the limit on the $z\sim 5$ QLF by
\citet{2012ApJ...756..160I} due to systematic errors in their
selection function\footnote{\citet{2012ApJ...756..160I} underestimated
  the dispersion in rest-frame UV quasar colours with respect to SDSS
  at all redshifts (their Figure~4). Contrary to their claim,
  photometric errors have a small effect on the colour distribution of
  SDSS quasars given the statistical errors of $<0.03$\,mag in $gri$
  for 90 per cent of the SDSS DR7 bright quasar sample ($i<19.1$) and
  a relative calibration error of $\sim 1$ per cent
  \citep{2008ApJ...674.1217P}.}.

@@

\begin{figure*}
  \begin{center}
    % rtg2.draw_g_paper() 
    \includegraphics[scale=0.65]{g.pdf}
  \end{center}
  \caption{AGN contribution to the hydrogen photoionisation rate,
    assuming unit escape fraction, when the AGN luminosity function is
    integrated down to $M_{1450}=-21$ (blue curve and shaded region)
    and $M_{1450}=-18$ (red curve and shaded region).  The shaded
    regions show the one-sigma (68.26\%) uncertainty.  Also shown are
    the photoionization rate measurements by \citet[filled
      circles]{2013MNRAS.436.1023B}, \citet[inverted
      triangles]{2011MNRAS.412.2543C},
    \citet[triangles]{2011MNRAS.412.1926W}, \citet[blue
      hexagons]{2018MNRAS.473..560D} and \citet[green
      squares]{2018ApJ...855..106D}, and
    \citet[pentagons]{2017MNRAS.467.3172G}, and models of
    \citet[dotted brown curve]{2012ApJ...746..125H}, the QSO
    contribution in this model (dashed grey), \citet[dashed
      brown]{2015ApJ...813L...8M}, the QSO contribution from the model
    of \citet[dashed orange]{2015MNRAS.451L..30K}, \citet[dotted
      grey]{2017ApJ...837..106O}, and \citet[dashed
      grey]{2018arXiv180104931P}.  The photoionisation rate derived
    from the luminosity function fits of \citet{2015AA...578A..83G}
    are shown by the red open circles.}
  \label{fig:gammapi}
\end{figure*}

@@

Our calculation of the AGN contribution to the UV background follows
previous work on UV background synthesis models
\citep[e.g.][]{1996ApJ...461...20H,2012ApJ...746..125H}.  The main
quantity of interest is the \ion{H}{i} photoionization rate of the UV
background
\begin{equation}
  \Gamma_\ion{H}{i}\left(z\right)=\int_{\nu_{912}}^\infty\mathrm{d}\nu
  \frac{4\pi J_\nu\left(\nu,z\right)}{h\nu} \sigma_\ion{H}{i}\left(\nu\right)\quad,
  \label{eqn:gammapi}
\end{equation}
where $h$ is Planck's constant, $\sigma_\ion{H}{i}(\nu)$ is the
\ion{H}{i} photoionization cross-section \citep{1997MNRAS.292...27H,
  2011piim.book.....D}, and
\begin{multline}
  J_\nu(\nu, z)=\frac{c}{4\pi}\int_{z}^\infty\mathrm{d}z^\prime\frac{\left(1+z\right)^3}{H\left(z^\prime\right)\left(1+z^\prime\right)}\epsilon_\nu\left(\nu_\mathrm{em},z^\prime\right)\\
  \times\exp{\left[-\tau_\mathrm{eff}\left(\nu, z, z^\prime\right)\right]}
  \label{eqn:flux}
\end{multline}
is the angle- and space-averaged specific intensity of the UV
background. In the above equation
$H\left(z^\prime\right)=H_0(\Omega_\mathrm{m}\left(1+z^\prime\right)^3+\Omega_\Lambda)^{1/2}$
is the Hubble parameter, and
$\epsilon_\nu\left(\nu_\mathrm{em},z^\prime\right)$ is the comoving
emissivity of all \ion{H}{i} Lyman continuum sources at redshift
$z^\prime>z$ and emitted frequency
$\nu_\mathrm{em}=\nu\left(1+z^\prime\right)/\left(1+z\right)>\nu_{912}$.
In practice, we adopt an upper limit $z^\prime_\mathrm{max}=15$ when
integrating Equation~\eqref{eqn:flux}.  Assuming that quasars are the
only ionizing sources and considering Equation~\eqref{eqn:sed} we have
\begin{equation}
  \epsilon_\nu\left(\nu_\mathrm{em},z^\prime\right) = \epsilon_{912}\left(z^\prime\right)\left(\frac{\nu_\mathrm{em}}{\nu_{912}}\right)^{-1.70}\quad,
  \label{eqn:epsilon_freq}
\end{equation}
with $\epsilon_{912}\left(z^\prime\right)$ being represented by
Equations~\eqref{eqn:e912_18} and \eqref{eqn:e912_21} for our two
magnitude limits, respectively.  For Poisson-distributed absorbers
with an \ion{H}{i} column density distribution
$f\left(N_\ion{H}{i},z^{\prime\prime}\right)=\partial^2n/\left(\partial
N_\ion{H}{i}\partial z^{\prime\prime}\right)$, the effective optical
depth to \ion{H}{i} Lyman continuum photons travelling between
redshifts $z^\prime$ and $z$ is \citep{1980ApJ...240..387P}
\begin{equation}
  \tau_\mathrm{eff}\left(\nu,z,z^\prime\right) = \int_{z}^{z^\prime}\mathrm{d}z^{\prime\prime}\int_0^\infty
  \mathrm{d}N_\ion{H}{i} f\left(N_\ion{H}{i},z^{\prime\prime}\right)\left[1-e^{-\tau_1}\right]\quad,
   \label{eqn:taueff}
\end{equation}
where $\tau_1\approx
N_\ion{H}{i}\sigma_\ion{H}{i}\nu\left(1+z^{\prime\prime}\right)/\left(1+z\right)$
is the Lyman continuum optical depth through an individual
absorber\footnote{The helium content of the absorber can be ignored
  due to the small \ion{H}{i} photoionization cross section at
  $\lambda<304$\,\AA.}.  For the \ion{H}{i} column density
distribution $f\left(N_\ion{H}{i},z\right)$ we adopt the piecewise
power-law parametrization by \citet{2012ApJ...746..125H} that is
consistent with $f\left(N_\ion{H}{i},z\right)$ measurements at
$z<3.5$, and roughly reproduces both the \ion{H}{i} Ly$\alpha$
effective optical depth \citep[but not in detail---see][]{2015MNRAS.450.4081P,2017MNRAS.464..897B,2017ApJ...837..106O}
and the measured mean free path to \ion{H}{i} Lyman limit photons at
$z<5.5$ \citep{2009ApJ...705L.113P,2014MNRAS.445.1745W}.
\citet{2018arXiv180109693K} suggest that variations in
$f\left(N_\ion{H}{i},z\right)$ result in modest (10--40 per cent)
changes in $\Gamma_\ion{H}{i}$ at $z<3$. We note, however, that at
$z\ga 3.5$ all UV background synthesis models are based on brazen
extrapolations of $f\left(N_\ion{H}{i},z\right)$, whose detailed shape
for (partial) Lyman limit systems is not well constrained at these
redshifts \citep{2010ApJ...718..392P}.

@@

\begin{figure*}
  \begin{center}
    % corner.corner() with bins.lfs[-4]
    \includegraphics[width=0.65\textwidth]{corner_z3p9.pdf}
  \end{center}
  \caption{Posterior distributions of the four double-power-law
    parameters in the $3.7\leq z < 4.1$ redshift bin.  The blue
    squares indicate median values.  Similar behaviour of the
    posterior distributions is seen in all other redshift bins defined
    in Section~\ref{sec:bins}.
    \label{fig:corner}}
\end{figure*}

@@

@@

  \begin{tabular}{ccc....}

@@

@@

    0.31$^a$ & 0.10 & 0.40 & 0.53^{+0.02}_{-0.02} & 0.70^{+0.03}_{-0.03} & 0.30^{+0.01}_{-0.01} & 0.40^{+0.01}_{-0.01} \\
    0.50$^a$ & 0.40 & 0.60 & 0.75^{+0.01}_{-0.01} & 1.00^{+0.02}_{-0.02} & 0.58^{+0.01}_{-0.01} & 0.78^{+0.01}_{-0.01} \\
    0.72 & 0.60 & 0.80 & 1.73^{+0.05}_{-0.05} & 2.30^{+0.07}_{-0.07} & 1.19^{+0.02}_{-0.02} & 1.57^{+0.02}_{-0.03} \\
    0.91 & 0.80 & 1.00 & 2.85^{+0.16}_{-0.16} & 3.81^{+0.21}_{-0.21} & 2.10^{+0.05}_{-0.05} & 2.78^{+0.07}_{-0.06} \\
    1.10 & 1.00 & 1.20 & 3.72^{+0.10}_{-0.10} & 4.94^{+0.13}_{-0.12} & 2.91^{+0.04}_{-0.04} & 3.87^{+0.05}_{-0.06} \\
